# How and when is care used by the dying?

The previous section showed which services people accessed. We now consider how much
healthcare is used. We consider this by proximity to death, by cause and by age.

## How much care and when; service use by proximity to death

```{r utilisation curve}
utilisation_curve_plot <- function(data, loess_span = 0.15) {
  data %>%
    ggplot(aes(proximity_to_death_days, stp)) +
    geom_smooth(span = loess_span,
                formula = y ~ x,
                method = "loess",
                fill = NA,
                colour = su_theme_cols()[["orange"]]) +
    geom_smooth(aes(y = region),
              span = loess_span,
              formula = y ~ x,
              method = "loess",
              fill = NA,
              colour = after_scale(alpha(su_theme_cols()[["charcoal"]], 0.4))) +
    geom_point(size = 0.2,
               shape = "circle filled",
               colour = su_theme_cols()[["orange"]],
               fill = after_scale(alpha(su_theme_cols()[["orange"]], 0.4))) +
    geom_vline(xintercept = 3*365/12,
               linetype = "dashed",
               colour = su_theme_cols()[["dark_charcoal"]]) +
    scale_x_proximity_to_death_days() +
    scale_y_continuous(labels = comma_format(scale = 1000, accuracy = 1),
                       position = "right") +
    theme(panel.grid.major = element_line(colour = "grey",
                                          size = 0.25,
                                          linetype = "dotted"),
          axis.text.y.left = element_blank(),
          axis.title.y.right = element_text(margin = margin(0, 0, 0, 10)),
          axis.ticks.y.left = element_blank(),
          legend.position = "bottom") +
    labs(x = "Proximity to Death (Months)",
         y = "Activity per 1,000 Decedants",
         colour = "") +
    annotate("text",
             x = 3*365/12+10,
             y = max(c(data$stp, data$region)),
             label = "3 months prior to death",
             hjust = 1,
             size = 3)
}

utilisation_plots <- list(
  stp = activity,
  region = activity_region
) %>%
  map_dfr(.id = "type", function(.x) {
    .x %>%
      mutate_at("pod_type",
                ~ifelse(pod_summary_group == "Critical Care Bed Day",
                        "Critical Care Bed Day",
                        as.character(.x))) %>%
      mutate_at("pod_type",
                fct_relevel,
                c(levels(activity$pod_type), "Critical Care Bed Day")) %>%
      count(pod_type, proximity_to_death_days)
  }) %>%
  inner_join(tribble(~type, ~p,
                     "stp", nrow(mpi),
                     "region", nrow(mpi_region)),
             by = "type") %>%
  mutate_at("n", ~n / p) %>%
  select(-p) %>%
  pivot_wider(names_from = type, values_from = n) %>%
  modify_at(vars(region, stp), replace_na, 0) %>%
  group_nest(pod_type) %>%
  mutate(plot = pmap(list(data = data,
                          loess_span = c(0.05, 0.15, 0.15, 0.05, 0.05)),
                     utilisation_curve_plot))
```

```{r utilisation curve grouped, message=FALSE,  echo=FALSE, results = "asis"}
utilisation_curve_plot_grouped <- function(pod_type, data, loess_span = 0.15) {
  data %>%
    ggplot(aes(proximity_to_death_days, n, colour = group)) +
    geom_smooth(span = loess_span,
                formula = y ~ x,
                method = "loess",
                size = 0.5,
                fill = NA) +
    geom_vline(xintercept = 3*365/12,
               linetype = "dashed",
               colour = su_theme_cols()[["dark_charcoal"]]) +
    scale_x_proximity_to_death_days() +
    scale_y_continuous(labels = comma_format(scale = 1000, accuracy = 1),
                       position = "right") +
    theme(panel.grid.major = element_line(colour = "grey",
                                          size = 0.25,
                                          linetype = "dotted"),
          axis.text.y.left = element_blank(),
          axis.title.y.right = element_text(margin = margin(0, 0, 0, 10)),
          axis.ticks.y.left = element_blank(),
          legend.position = "bottom") +
    labs(x = "Proximity to Death (Months)",
         y = "Activity per 1,000 Decedants",
         colour = "") +
    annotate("text",
             x = 3*365/12+10,
             y = max(data$n),
             label = "3 months prior to death",
             hjust = 1,
             size = 3)
}

utilisation_plots_grouped <- activity %>%
  mutate_at("pod_type",
            ~ifelse(pod_summary_group == "Critical Care Bed Day",
                    "Critical Care Bed Day",
                    as.character(.x))) %>%
  mutate_at("pod_type",
            fct_relevel,
            c(levels(activity$pod_type), "Critical Care Bed Day")) %>%
  inner_join(select(mpi, su_pat_id, group), by = "su_pat_id") %>%
  filter(group != "(Missing)") %>%
  count(pod_type, group, proximity_to_death_days) %>%

  inner_join(count(mpi, group, name = "p"), by = "group") %>%
  mutate_at("n", ~n/p) %>%
  
  group_nest(pod_type) %>%
  mutate(plot = pmap(list(data = data,
                          loess_span = c(0.05, 0.15, 0.15, 0.05, 0.05)),
                     utilisation_curve_plot_grouped))
```

In this sub-section we consider how use of care changes over time, as people approach death. For
each type of activity, we show how service use increases, decreases and at what point a peak is
reached. Time here means ‘time before death’, regardless of any actual calendar date. Activity
taking place on the day of death is a time of 0 days, taking place the day before death is 1 day etc.
In the two year period prior to death we show rates of service use for those dying in the `r stp_name`
(yellow dots); the utilisation curve for the `r stp_name` (yellow line); and the utilisation curve
for the `r region_name` as a whole (dashed grey line). 

### Urgent care use rises rapidly and peaks on day of death

Urgent service events start low and increase slowly for much of the period until a few months prior
to death when there is a rapid rate of increase, rising to a sharp peak at the point of death. The rate
and pattern of service use in the `r stp_name` STP is the same as the `r region_name`.

#### Figure 20 : Urgent service events per 1,000 decedents by proximity to death in days - `r stp_name` (yellow dots and line) relative to `r region_name` region (grey line)

```{r utilisation_plots - urgent service events}
utilisation_plots$plot[[which(utilisation_plots$pod_type == "Urgent Service Event")]]
```

### Use of planned contacts increase steadily, peaking in weeks before death

Planned contacts rise steadily throughout the period until a sharp peak in the weeks prior to death,
at which point they decline. Although the pattern of service use is similar there is a consistently
higher rate in the `r stp_name` when compared to the `r region_name`.

#### Figure 21 : Planned contacts per 1,000 decedents by proximity to death in days - `r stp_name`. STP (yellow dots and line) relative to `r region_name` region (grey line)

```{r utilisation_plots - planned contacts}
utilisation_plots$plot[[which(utilisation_plots$pod_type == "Planned Contact")]]
```

### Use of planned admissions increases steadily, peaking in months before death

Planned admissions also rise steadily throughout the period with a rounded peak a matter of
months prior to death, at which point they decline. There is a higher rate of utilisation in the 
`r stp_name` further away from death. In the last year of life rates start to
converge and eventually drop lower those seen in the `r region_name`.

#### Figure 22 : Planned admissions per 1,000 decedents by proximity to death in days - `r stp_name` STP (yellow dots and line) relative to `r region_name` region (grey line)

```{r utilisation_plots - planned admissions}
utilisation_plots$plot[[which(utilisation_plots$pod_type == "Planned Admission")]]
```

## Use of hospital beds rises rapidly and peaks on day of death

Use of bed days, created when an admission to a bed takes place, closely follows the pattern seen
in urgent service events. This is because, as seen in analysis of patterns of service use (in subsection 5.4), bed occupancy for the decedent population usually follows an emergency admission.
Therefore, these patterns of service use are intrinsically linked. The rate and pattern of service use
in the `r stp_name` is the same as the `r region_name`.

#### Figure 23 : Bed days per 1,000 decedents by proximity to death in days - `r stp_name` STP (yellow dots and line) relative to `r region_name` region (grey line)

```{r utilisation_plots - bed days}
utilisation_plots$plot[[which(utilisation_plots$pod_type == "Bed")]]
```

## Critical care days are concentrated in the last month of life

Again, this closely follows the pattern seen in urgent service events. The rate and pattern of service
use in the `r stp_name` is the same as the `r region_name`.

#### Figure 24 : Critical care days per 1,000 decedents by proximity to death in days - `r stp_name` STP (yellow dots and line) relative to `r region_name` region (grey line)

```{r utilisation_plots - critical care bed days}
utilisation_plots$plot[[which(utilisation_plots$pod_type == "Critical Care Bed Day")]]
```

### Use of urgent service events show similar patterns by cause

Utilisation curves by cause show patterns which are very similar for urgent service events (Figure
25). Where curves do diverge, this occurs in the last 3 months of life.

#### Figure 25 : Urgent service events per 1,000 by cause and proximity to death in days - `r region_name` STP

```{r utilisation_plots_grouped - urgent service events}
utilisation_plots_grouped$plot[[which(utilisation_plots_grouped$pod_type == "Urgent Service Event")]]
```

### Use of planned contacts and admissions is much higher for those dying from cancer

Utilisation curves by cause of death in Figure 26 show that those dying from cancer dominate use
of planned contacts. Whilst only a fifth of deaths are due to cancer we have seen that people in this
group experience a relatively high volume of planned contacts (sub-section 5.4.2).

Cancer patients will experience a high volume of planned care due to cancer treatment regimens.
These regimens require a regular sequence of multiple hospital visits, leading to the high volumes
of planned care for cancer patients. Many planned contacts are outpatient attendances, a setting
used for delivering radiotherapy regimens.

#### Figure 26 : Planned contacts per 1,000 by cause and proximity to death in days - `r stp_name` STP

```{r utilisation_plots_grouped - planned contact}
utilisation_plots_grouped$plot[[which(utilisation_plots_grouped$pod_type == "Planned Contact")]]
```

Utilisation curves by cause of death in Figure 27 show that those dying from cancer also dominate
the use of planned admissions. We have seen that people in this group also experience a relatively
high volume of planned admissions (again, in sub-section 5.4.2).

As mentioned in above, cancer patients will experience a high volume of planned care due to
cancer treatment regimens. These regimens require a regular sequence of multiple hospital visits,
leading to the high volumes of planned care for cancer patients. Many planned admissions are
daycases, a setting used for delivering chemotherapy regimens.

#### Figure 27 : Planned admissions per 1,000 by cause and proximity to death in days - `r stp_name` STP

```{r utilisation_plots_grouped - planned admissions}
utilisation_plots_grouped$plot[[which(utilisation_plots_grouped$pod_type == "Planned Admission")]]
```

### Use of bed days – and critical care bed days - show similar patterns by cause

Utilisation curves by cause show patterns which are very similar for bed days (Figure 28). Where
curves do diverge, this occurs in the last 6 months of life.

#### Figure 28 : Bed days per 1,000 by cause and proximity to death in days - `r stp_name` STP

```{r utilisation_plots_grouped - bed}
utilisation_plots_grouped$plot[[which(utilisation_plots_grouped$pod_type == "Bed")]]
```

Utilisation curves by cause also show patterns which are very similar for critical care bed days
(Figure 29). Where curves do diverge, this occurs in the last 3 months of life.

#### Figure 29 : Critical care days per 1,000 by cause and proximity to death in days - `r stp_name` STP

```{r utilisation_plots_grouped - critical care bed days}
utilisation_plots_grouped$plot[[which(utilisation_plots_grouped$pod_type == "Critical Care Bed Day")]]
```

## Does age at death influence service use?

In this sub-section we consider the use of care by age at death. We do this for each service
within an activity type.

We consider this for two periods:

- the final year of life (0-12 months before death); and,
- the penultimate year of life (12 – 24 months before death).

```{r utilisation rates}
utilisation_rates <- list(stp = activity,
     region = activity_region) %>%
  map_dfr(.id = "type", function(.x) {
    .x %>%
      inner_join(select(mpi_region, su_pat_id, age),
                 by = "su_pat_id") %>%
    mutate(activity_year = as.numeric(proximity_to_death_days >= 365)) %>%
    mutate_at("age", ~case_when(.x > 90 ~ 90,
                                .x < 50 ~ 50,
                                TRUE ~ .x)) %>%
    mutate_if(is.factor, as.character) %>%
    count(pod_type, pod_summary_group, activity_year, age,
          name = "activity")
  }) %>%
  arrange(type, pod_type, pod_summary_group, activity_year, age) %>%
  complete(nesting(pod_type, pod_summary_group),
           activity_year = 0:1,
           age = seq(50, 90),
           fill = list(activity = 0)) %>%
  inner_join(
    bind_rows(
      mpi %>%
        mutate_at("age",
                  ~case_when(.x > 90 ~ 90,
                             .x < 50 ~ 50,
                             TRUE ~ .x)) %>%
        count(age, name = "deaths") %>%
        mutate(type = "stp"),
      mpi_region %>%
        mutate_at("age",
                  ~case_when(.x > 90 ~ 90,
                             .x < 50 ~ 50,
                             TRUE ~ .x)) %>%
        count(age, name = "deaths") %>%
        mutate(type = "region")
    ),
    by = c("type", "age")
  ) %>%
  mutate(utilisation = activity / deaths) %>%
  select(-activity, -deaths) %>%
  pivot_wider(names_from = "type", values_from = "utilisation") %>%
  mutate_at("pod_summary_group", ~fct_reorder(.x, region, .fun = sum)) %>%
  pivot_longer(c(region, stp), names_to = "type", values_to = "value") %>%
  group_nest(pod_type, type) %>%
  mutate(plot = map(data, function(data) {
    data %>%
      mutate_at("pod_summary_group", fct_drop) %>%
      group_by(pod_summary_group) %>%
      # make sure we have a sensible number of rows per pod_summary_group, we should have 82 rows
      # 2*(90-50+1)
      filter(sum(is.na(value)) < 82*.75) %>%
      ungroup() %>%
      mutate_at("activity_year",
                ifelse,
                "Penultimate year of life (12-24 months before death)",
                "Final year of life (0-12 months before death)") %>%
      ggplot(aes(age, value, colour = activity_year)) +
      geom_smooth(fill = NA, method = "loess", formula = y ~ x, na.rm = TRUE,
                  lwd = 0.75) +
      geom_point(aes(fill = after_scale(alpha(colour, 0.4))),
                 shape = "circle filled",
                 na.rm = TRUE) +
      expand_limits(y = 0) +
      facet_wrap(vars(fct_rev(pod_summary_group)), drop = FALSE) +
      theme(legend.position = "bottom") +
      labs(colour = "",
           x = "Age of decedent",
           y = "Average use")
  }))
```

### Use of urgent service events is not influenced by age at death

Age specific use of urgent service events in Figure 30 shows that:

- Largest areas of use are emergency admissions and A&E attendances. Use of these services
does not differ greatly by age at death. Use is however much higher in the final year of life than
in the penultimate year of life. This signifies that what drives utilisation of these services is not
age, but proximity to death;
- Calls to 111 are used less often than the other urgent services. For those dying aged 70 and
over, as age increases so does use of 111. Use is again higher in the final year of life; and,
- There is less use of 111 compared to the `r region_name` (Figure 31). Age patterns are similar.

#### Figure 30 : Urgent service event use by age at death in final and penultimate year of life - `r stp_name` STP

```{r utilisation rates - Urgent Service Events (stp)}
filter(utilisation_rates, pod_type == "Urgent Service Event", type == "stp")$plot[[1]]
```

#### Figure 31 : Urgent service event use by age at death in final and penultimate year of life - `r region_name` region

```{r utilisation rates - Urgent Service Events (region)}
filter(utilisation_rates, pod_type == "Urgent Service Event", type == "region")$plot[[1]]
```

### Use of planned contacts decreases for older decedents

Age specific use of planned contacts in Figure 32 shows that:

- Largest use is for outpatient attendances. As age at death increases use of outpatient
attendances decrease;
- Use of outpatient attendances is higher in final year of life than in the penultimate year.
However, the difference between years diminishes as age increases;
- There is less use of mental health compared to the `r region_name` (Figure 33). Age patterns are
similar.

#### Figure 32 : Planned contact use by age at death in final and penultimate year of life - `r stp_name` STP

```{r utilisation rates - Planned Contacts (stp)}
filter(utilisation_rates, pod_type == "Planned Contact", type == "stp")$plot[[1]]
```

#### Figure 33 : Planned contact use by age at death in final and penultimate year of life – `r region_name` region

```{r utilisation rates - Planned Contacts (region)}
filter(utilisation_rates, pod_type == "Planned Contact", type == "region")$plot[[1]]
```

### Use of planned admissions also decreases for older decedents

Age specific use of planned admissions in Figure 34 shows that:

- Largest areas of use are day case admissions and regular attendances. Generally, as age at
death increases use of these services decreases;
- There is more use of regular attendances compared to the `r region_name` (Figure 35). Age patterns
are similar.

#### Figure 34 : Planned admission use by age at death in final and penultimate year of life - `r stp_name` STP

```{r utilisation rates - Planned Admissions (stp)}
filter(utilisation_rates, pod_type == "Planned Admission", type == "stp")$plot[[1]]
```

#### Figure 35 : Planned admission use by age at death in final and penultimate year of life - `r region_name` region

```{r utilisation rates - Planned Admissions (region)}
filter(utilisation_rates, pod_type == "Planned Admission", type == "region")$plot[[1]]
```

### Older decedents stay in hospital longer

Age specific use of bed days in Figure 36 shows that:

- Use of bed days is dominated by emergency admission bed stays. Use of emergency bed days
increases as age at death increases. We have seen in earlier sub-section (6.2.1) that age does
not influence use of emergency admissions. However, we see here that age does influence
length of stay. And that when admitted, the older the decedent the longer the length of stay;
- In the final year of life, the oldest decedents spend approximately an additional seven days in
hospital than the youngest decedents; and,
- Volume and age patterns of use are similar to the `r region_name` (Figure 37).

#### Figure 36 : Bed day use by age at death in final and penultimate year of life - `r stp_name`

```{r utilisation rates - Bed Days (stp)}
filter(utilisation_rates, pod_type == "Bed", type == "stp")$plot[[1]]
```

#### Figure 37 : Bed day use by age at death in final and penultimate year of life - `r region_name`

```{r utilisation rates - Bed Days (region)}
filter(utilisation_rates, pod_type == "Bed", type == "region")$plot[[1]]
```
