# How many deaths have there been? How many will there be?

Trends in the number of expected deaths and age at death are driven by broader demographic
changes, including the gains in life expectancy seen throughout the 20th century. While dramatic in
a broader historical context, these changes have unfolded slowly over decades. It is possible to
predict and plan for deaths in the local population. This section therefore provides a basic scaling
of historic and forecast numbers of deaths by gender and age group. 

##	Having declined for decades, deaths are set to increase

The population of England has grown almost every year since the end of the Second World War.
Figure 1 shows that until the late 1970s the number of deaths per annum also grew although at a
slower rate than the population. Since the early 1980s the number of deaths per year has fallen and
the number of deaths in 2009 were the lowest that had been seen since 1952. This continued until
2010 when the trend reversed sharply.

#### Figure 1: Deaths in England, long term trends and forecasts

*[Insert figure]*

## Deaths will increase for both males and females

Trends and forecasts in `r stp_name` reflect those in England. Figure 2 shows the annual increase in deaths over the last decade. In 2018/19 `r comma(stp_deaths_18)` adults died in `r stp_name`. Between 2020 and 2030 the number of deaths is expected to grow `r percent(stp_deaths_pcnt)` to `r comma(stp_deaths_30)` per annum. As the size of the decedent population grows so too will demand on services.

There has also been a shift towards increasing numbers of deaths in males, narrowing the gap between genders. In future years deaths for males are predicted to be consistently higher than deaths for females.

#### Figure 2: Historical and forecast deaths by gender - `r stp_name`

```{r historic and forecast deaths}
local({
  df <- list(
    historical = historical_deaths %>%
      select(-stp) %>%
      mutate(point_val = deaths),
    forecast = forecast_deaths %>%
      group_by(year, sex) %>%
      summarise_at("est_deaths", sum) %>%
      rename(deaths = est_deaths) %>%
      filter(year > 2019)
  ) %>%
    bind_rows(.id = "type") %>%
    mutate_at("type", fct_rev)
  
  df %>%
    ggplot(aes(year, deaths, colour = sex)) +
    geom_line(aes(linetype = type)) +
    geom_point(aes(y = point_val), na.rm = TRUE, show.legend = FALSE) +
    geom_vline(xintercept = 2019,
               colour = su_theme_cols("grey"),
               linetype = "dotted") +
    scale_x_continuous(breaks = seq(2010, 2040, 5)) +
    annotate("text",
             x = min(df$year),
             y = max(df$deaths),
             label = "Historical",
             hjust = 0) +
    annotate("text",
             x = 2020,
             y = max(df$deaths),
             label = "Forecast",
             hjust = 0) +
    scale_y_continuous(labels = comma_format(accuracy = 1)) +
    scale_linetype_manual(values = c("forecast" = "dashed",
                                     "historical" = "solid"),
                          guide = "none") +
    labs(colour = "gender", x = "", y = "") +
    theme(legend.position = c(1, 0),
          legend.justification = c(1, 0),
          legend.background = element_blank(),
          legend.key = element_blank(),
          legend.title = element_blank())
})
```

## The oldest decedents account for most of the increase

Figure 3 shows forecast deaths by age group. The greatest number of deaths is in those aged 85+.
This is also the age group with the most significant expected increase. The needs of these older
decedents will therefore have a greater impact on future demand.

#### Figure 3 : Forecast deaths by age group - `r stp_name`

```{r forecast deaths}
forecast_deaths %>%
  mutate_at("age_group",
            cut,
            breaks = c(0,18,65,75,85,Inf),
            labels = c("0-17", "18-64","65-74","75-85","85+"),
            right = FALSE) %>%
  arrange(age_group, year) %>%
  group_by(age_group, year) %>%
  filter(year >= 2020) %>%
  summarise_at("est_deaths", sum) %>%
  mutate(first_year = ifelse(year == min(year),
                             comma(est_deaths, accuracy = 1),
                             NA),
         last_year  = ifelse(year == max(year),
                             paste0(comma(est_deaths, accuracy = 1),
                                    " (",
                                    percent(est_deaths / first(est_deaths) - 1,
                                           accuracy = 1),
                                   ")"),
                             NA)) %>%
  ungroup() %>%
  ggplot(aes(year, est_deaths, colour = age_group)) +
  geom_line(linetype = "dashed") +
  geom_text_repel(aes(year-1, est_deaths, label = first_year),
                  na.rm = TRUE,
                  direction = "y",
                  hjust = 1,
                  size = 3,
                  segment.colour = NA,
                  point.padding = NA,
                  show.legend = FALSE) +
  geom_text_repel(aes(year+1, est_deaths, label = last_year),
                  na.rm = TRUE,
                  direction = "y",
                  hjust = 0,
                  size = 3,
                  segment.colour = NA,
                  point.padding = NA,
                  show.legend = FALSE) +
  labs(colour = "",
       x = "",
       y = "") +
  scale_x_continuous(breaks = seq(2020, 2040, 5),
                     expand = expansion(c(0.10, 0.15))) +
  scale_colour_manual(values = su_theme_cols("slate",
                                             "orange",
                                             "charcoal",
                                             "blue",
                                             "red") %>% unname()) +
  scale_y_continuous(labels = comma) +
  theme(legend.position = "bottom",
        legend.key = element_blank())
``` 
