# Where do people die?

The previous section described historic and forecast numbers of deaths. Yet *where* people die is
important to the quality and experience of their death. This section therefore examines place of
death. It begins by considering how place of death varies by cause, before moving on to consider
differences by deprivation and by gender. It also considers the length of stay in hospital for those
who die in hospital. Concluding by exploring how place of death is a factor in the level of
unrelieved pain for palliative patients, calculating the number of palliative patients dying with their
pain not fully relieved.

## How place and cause of death are recorded and assigned

When a person dies a doctor involved in their care completes a medical death certificate which is
used to formally register the death. This contains detailed information about the individual, their
place of death and underlying cause of death.

For the purposes of this report place of death is assigned to one of the five categories defined by
the National End of Life Care Intelligence Network. They are:
- Home;
- Care home;
- Hospice;
- Hospital; and,
- Elsewhere.

The underlying cause of death is assigned to one of the five cause groups below. These groups are
based on research by Dr June Lunney and Dr Joanne Lynn

- Cancer;
- Frailty;
- Organ failure;
- Sudden death; and,
- Other terminal illness.

It can however be difficult to assign deceased patients to the frailty group based on cause of death
alone. To define frailty we therefore utilise work done by Whole Systems Partnership for the
National End of Life Care Intelligence Network (NEoLCIN). This work additionally assigns patients
by age groups on the following basis:

- aged 65-74 then 10% of deaths are frailty related;
- aged 75-84 then 30% of deaths are frailty related; and,
- aged 85+ then 80% of deaths are frailty related.  

## Fewer die at home than would like to

```{r 4.2 figures}
die_home_pcnt <- nrow(filter(mpi, location_type == "Home")) / stp_deaths_18
```
Although 66% of people say they would like to die at home, just `r percent(die_home_pcnt)` of people
in `r stp_name` do so.

Figure 4 shows that place of death differs significantly by cause, in that:

- Frailty is the single largest underlying cause of death, accounting for close to half of all deaths.
Frailty has the largest proportion of deaths in a care home setting;
- Cancer is the cause of death for around a fifth of the population. 22% of cancer patients die in
a hospice setting. This is considerably higher than other causes;
- Organ failure and sudden death have a large proportion of deaths in a hospital setting
although there are still a substantial number of deaths occurring at home for both these
groups; and,
- Other terminal illness (OTI) represent the smallest cause of death group within the population,
but this category has a large proportion of deaths in a hospital setting.

If hospice care for organ failure patients could be organised along similar lines as for cancer care
an additional 252 people could die in a hospice rather than a hospital.

When compared to the `r region_name` (Figure 5) profiles by cause are generally similar. However, in
`r stp_name` there are a lower proportion of deaths in hospital for all causes.

#### Figure 4 : Proportion of decedent population by cause and place of death - `r stp_name`

```{r place of death by cause of death mekko (stp)}
mpi %>%
  filter(location_type != "Unknown",
         group != "(Missing)") %>%
  mutate_at(vars(group),
            compose(fct_rev, fct_recode),
            "Sudden\nDeath" = "Sudden Death",
            "OTI" = "Other Terminal Illness") %>%
  mekko_chart(group, location_type) +
  place_of_death_fill() 
```

#### Figure 5 : Proportion of decedent population by cause and place of death - `r region_name` region

```{r place of death by cause of death mekko (region)}
mpi_region %>%
  filter(location_type != "Unknown",
         group != "(Missing)") %>%
  mutate_at(vars(group),
            compose(fct_rev, fct_recode),
            "Sudden\nDeath" = "Sudden Death",
            "OTI" = "Other Terminal Illness") %>%
  mekko_chart(group, location_type) +
  place_of_death_fill() 
```

## People from deprived areas are more likely to die in hospital

Figure 6 shows that for decedents living in more deprived areas there were a higher proportion of
deaths in hospital. This is offset by smaller proportion of deaths taking place in care homes for
these decedents. Proportions differ but this is the same profile seen for the `r region_name` in Figure 7.

#### Figure 6 : Proportion of decedent population by deprivation quintile and place of death - `r stp_name`

```{r place of death by deprivation mekko (stp)}
mpi %>%
  filter(location_type != "Unknown",
         !is.na(imd_quintile)) %>%
  mutate(across(imd_quintile, as.factor)) %>%
  mekko_chart(imd_quintile, location_type,
              x_labels = c("Most Deprived", "Least Deprived"),
              x_breaks = c(0.075, 0.925)) +
  place_of_death_fill()
```

#### Figure 7 : Proportion of decedent population by deprivation quintile and place of death - `r region_name` region

```{r place of death by deprivation mekko (region)}
mpi_region %>%
  filter(location_type != "Unknown",
         !is.na(imd_quintile)) %>%
  mutate(across(imd_quintile, as.factor)) %>%
  mekko_chart(imd_quintile, location_type,
              x_labels = c("Most Deprived", "Least Deprived"),
              x_breaks = c(0.075, 0.925)) +
  place_of_death_fill()
```

## Fewer male deaths take place in care homes

Comparing between genders in Figure 8 there were a higher proportion of males dying either in
hospital or at home. The proportion of males dying in a care home was much lower when
compared to females. Proportions differ but again this is the same profile seen for the `r region_name` in
Figure 9.

#### Figure 8 : Proportion of decedent population by gender and place of death - `r stp_name`

```{r place of death by gender (stp)}
mpi %>%
  filter(location_type != "Unknown") %>%
  mekko_chart(sex, location_type) +
  place_of_death_fill()
```

#### Figure 9 : Proportion of decedent population by gender and place of death - `r region_name` region

```{r place of death by gender (region)}
mpi_region %>%
  filter(location_type != "Unknown") %>%
  mekko_chart(sex, location_type) +
  place_of_death_fill()
```

## For deaths in hospital length of stay is often short

```{r 4.5 figures}
die_hosp_pcnt_stp <- activity %>%
  inner_join(mpi, by = "su_pat_id") %>%
  filter(pod_summary_group == "Emergency Admission", discharge_date == dod) %>%
  nrow() / nrow(mpi)
die_hosp_pcnt_region <- activity_region %>%
  inner_join(mpi_region, by = "su_pat_id") %>%
  filter(pod_summary_group == "Emergency Admission", discharge_date == dod) %>%
  nrow() / nrow(mpi_region)
```

Of all decedents in `r stp_name` `r percent(die_hosp_pcnt_stp, accuracy = 1)` die in hospital after being admitted as an
emergency. For the `r region_name` the same figure is `r percent(die_hosp_pcnt_region, accuracy = 1)`.
Figure 10 shows, for entire decedent populations, the proportions by terminal episode length of stay.
Proportions for `r stp_name` are shown as bars. Proportions for the `r region_name` are shown as dots.
The length of stay for a terminal episode is often short. In `r stp_name`, as in the `r region_name`,
there are a higher proportion of decedents with the shortest stays (terminal episodes of 6 days or less).
Proportions are similar compared to the `r region_name`.

#### Figure 10 : Proportion of decedent population by emergency terminal episode length of stay - `r stp_name` indicated by bars, `r region_name` region indicated by dots

```{r tlos count plot}
tlos <- list(
  stp = activity,
  region = activity_region
) %>%
  map(semi_join, mpi_region, by = c("su_pat_id", "discharge_date" = "dod")) %>%
  map(filter, pod_summary_group == "Emergency Admission") %>%
  map(group_by, pod_summary_group) %>%
  map_dfr(count, proximity_to_death_days, .id = "type") %>%
  pivot_wider(names_from = type, values_from = n) %>%
  mutate(across(stp, replace_na, 0))

tlos %>%
  filter(proximity_to_death_days <= 42) %>%
  mutate(across(stp, `/`, nrow(mpi)),
         across(region, `/`, nrow(mpi_region))) %>%
  ggplot(aes(proximity_to_death_days, stp)) +
  geom_col(colour = su_theme_cols("orange"),
           fill = after_scale(alpha(su_theme_cols("orange"), 0.4)),
           width = 1) +
  geom_point(aes(y = region),
             colour = su_theme_cols("charcoal"),
             fill = after_scale(alpha(su_theme_cols("charcoal"), 0.4)),
             shape = "circle filled") +
  scale_x_continuous(breaks = seq(0, 70, by = 7)) +
  scale_y_continuous(expand = expansion(c(0, 0.05)),
                     labels = percent_format(accuracy = 1)) +
  labs(x = "Bed days for Terminal Spell",
       y = "% of Decedents")
```

## Place of death influences the likelihood of experiencing pain

Previous research has found a relationship between place of death and experience of pain11. People
who die in a hospice report the lowest level of pain at end of life compared to other settings.
Applying research from the Office of Health Economics12 to those receiving, or in need of, palliative
care in Coventry and Warwickshire (Table 1 and Table 2) suggests that in total a third of palliative
patients (2,048 people - using central estimate) may have died with their pain uncontrolled.
From these estimates; rates of unrelieved pain are highest for those palliative patients who die at
home; they are lowest for those who die in a hospice setting. This represents an opportunity to
explore the local picture of palliative care service provision and patient end of life experience to
ensure interventions to maximise pain control for people wherever they die.

#### Table 1 : By setting, the number of people at end of life whose pain is not relieved

*[insert table 1*

#### Table 2 : By setting, the number of people at end of life whose pain is only partially relieved

*[insert table 2]*
