# Which services are accessed before death?

In the two years before they die most people access some form of healthcare. In this section we
describe these patterns of use and how they vary by cause of death.
The datasets used in this section have been grouped into four different healthcare activity types:

|  ACTIVITY TYPE        | SERVICE GROUP                  |
|-----------------------|--------------------------------|
| Urgent service event  | Emergency Admissions           |
|                       | A&E Attendances                |
|                       | Calls to 111                   |
| Planned contact       | Planned Outpatient Attendances |
|                       | Mental Health Contact          |
|                       | IAPT Appointments              |
| Planned admission     | Daycases                       |
|                       | Elective Admissions            |
|                       | Regular Day/Night Admissions   |
| Bed days              | Critical Care                  |
|                       | All Other Bed Types            |

## Nearly all decedents access urgent care

Figure 11 shows that 9 in 10 people dying in `r stp_name` access urgent care
services in the two years prior to death. A similar but slightly smaller proportion also access
planned care. Access is lower for planned admissions (4 in 10) and much lower for critical care (1 in
10).

Access levels for `r stp_name` are shown as bars. Access levels for the `r region_name` are
shown as dots. Proportions are similar compared to the `r region_name`.

#### Figure 11 : Proportion of decedent population accessing healthcare activity types in two years prior to death - `r stp_name` STP indicated by bars (with percentage at bottom), `r region_name` region indicated by dots

```{r activity pcnt}
list(
  stp = activity,
  region = activity_region
) %>%
  map_dfr(.id = "type", function(.x) {
    .x %>%
      mutate_at("pod_type",
                ~ifelse(pod_summary_group == "Critical Care Bed Day",
                        "Critical Care Bed Day",
                        as.character(.x))) %>%
      group_by(pod_type) %>%
      distinct(su_pat_id) %>%
      count() %>%
      ungroup() %>%
      mutate_at("pod_type",
                fct_relevel,
                c(levels(activity$pod_type), "Critical Care Bed Day"))
  }) %>%
  inner_join(
    bind_rows(
      mpi %>%
        summarise(patients = n_distinct(su_pat_id)) %>%
        mutate(type = "stp"),
      
      mpi_region %>%
        summarise(patients = n_distinct(su_pat_id)) %>%
        mutate(type = "region"),
    ),
    by = c("type")
  ) %>%
  mutate_at("n", ~.x / patients) %>%
  select(-patients) %>%
  pivot_wider(names_from = "type", values_from = "n") %>%
  filter(pod_type != "Bed") %>%
  ggplot(aes(pod_type, stp)) +
  geom_col(aes(colour = pod_type,
               fill = after_scale(alpha(colour, 0.4)))) +
  geom_point(aes(y = region,
                 colour = pod_type,
                 fill = after_scale(alpha(colour, 0.4))),
             show.legend = FALSE,
             size = 2,
             shape = "circle filled",
             stroke = 1.5) +
  geom_text(aes(y = 0, label = percent(stp, accuracy = 1)),
            size = 3.5,
            vjust = 1.35) +
  scale_y_continuous(labels = percent,
                     expand = expansion(c(0.06, 0.05))) +
  expand_limits(y = c(0, 1)) +
  labs(x = "", y = "", colour = "", fill = "")  +
  theme(legend.position = "none",
        axis.line.x = element_blank(),
        axis.ticks.x = element_blank())
```

## Most decedents access emergency admissions, A&E and outpatient clinics; few access mental health services 

The proportion of people accessing different types of care in the last two years of life is shown in
Figure 12. It highlights radical differences by service type:

*[insert summary bullet points]*

#### Figure 12 : Proportion of decedent population accessing healthcare service types in two years prior to death - `r stp_name` STP indicated by bars (with percentage at bottom), `r region_name` region indicated by dots

```{r activity pcnt pod summary group}
list(
  stp = activity,
  region = activity_region
) %>%
  map_dfr(.id = "type", function(.x) {
    .x %>%
      group_by(pod_type, pod_summary_group, cc_pod_summary_group) %>%
      distinct(su_pat_id) %>%
      count() %>%
      ungroup() %>%
      mutate_at("pod_type",
                ~ifelse(pod_summary_group == "Critical Care Bed Day",
                        "Critical Care Bed Day",
                        as.character(.x))) %>%
      mutate_at("pod_summary_group",
                ~ifelse(pod_type == "Critical Care Bed Day",
                        paste0("Critical Care (",
                               word(cc_pod_summary_group, 1),
                               ")"),
                        as.character(pod_summary_group))) %>%
      mutate_at("pod_summary_group",
                fct_relevel,
                levels(activity$pod_summary_group) %>%
                  subset(., . != "Critical Care Bed Day"),
                "Critical Care (Emergency)",
                "Critical Care (Elective)")
  }) %>%
  inner_join(
    bind_rows(
      mpi %>%
        summarise(patients = n_distinct(su_pat_id)) %>%
        mutate(type = "stp"),
      
      mpi_region %>%
        summarise(patients = n_distinct(su_pat_id)) %>%
        mutate(type = "region"),
    ),
    by = c("type")
  ) %>%
  mutate_at("n", ~.x / patients) %>%
  select(-patients) %>%
  pivot_wider(names_from = "type", values_from = "n") %>%
  filter(pod_type != "Bed") %>% 
  ungroup() %>%
  mutate_at("pod_type",
            fct_relevel,
            c(levels(activity$pod_type) %>% subset(., . != "Bed"),
              "Critical Care Bed Day")) %>%
  ggplot(aes(pod_summary_group, stp)) +
  geom_col(aes(colour = pod_type,
               fill = after_scale(alpha(colour, 0.4)))) +
  geom_point(aes(y = region,
                 colour = pod_type,
                 fill = after_scale(alpha(colour, 0.4))),
             show.legend = FALSE,
             size = 2,
             shape = "circle filled",
             stroke = 1.5) +
  geom_text(aes(y = 0, label = percent(stp, accuracy = 1)),
            size = 3.5,
            vjust = 1.35) +
  scale_x_discrete(labels = partial(str_wrap, width = 10)) +
  scale_y_continuous(labels = percent,
                     expand = expansion(c(0.06, 0.05))) +
  expand_limits(y = c(0, 1)) +
  labs(x = "", y = "", colour = "", fill = "")  +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 30, hjust = 1),
        axis.line.x = element_blank(),
        axis.ticks.x = element_blank())
```

## Most cancer patients access planned admissions

Figure 13 looks at access to different service types by cause. This shows that:

- People dying from cancer access all types of service (except critical care) more than those
dying of other causes. Their access to planned admissions is significantly higher;
- Critical care access is much lower for people dying from frailty than from other causes; and,
- Access to urgent service events and planned contacts are broadly similar between causes of
death.

#### Figure 13 : Proportion of decedent population accessing healthcare activity types by cause in two years prior to death - `r stp_name` STP indicated by bars, `r region_name` region indicated by dots

```{r activity pcnt cod}
list(
  stp = activity %>%
    inner_join(select(mpi, su_pat_id, group), by = "su_pat_id"),
  region = activity_region %>%
    inner_join(select(mpi_region, su_pat_id, group), by = "su_pat_id")
) %>%
  map_dfr(.id = "type", function(.x) {
    .x %>%
      mutate_at("pod_type",
                ~ifelse(pod_summary_group == "Critical Care Bed Day",
                        "Critical Care Bed Day",
                        as.character(.x))) %>%
      group_by(pod_type, group) %>%
      distinct(su_pat_id) %>%
      count() %>%
      ungroup() %>%
      mutate_at("pod_type",
                fct_relevel,
                c(levels(activity$pod_type), "Critical Care Bed Day"))
  }) %>%
  inner_join(
    bind_rows(
      mpi %>%
        group_by(group) %>%
        summarise(patients = n_distinct(su_pat_id), .groups = "drop") %>%
        mutate(type = "stp"),
      
      mpi_region %>%
        group_by(group) %>%
        summarise(patients = n_distinct(su_pat_id), .groups = "drop") %>%
        mutate(type = "region"),
    ),
    by = c("type", "group")
  ) %>%
  mutate_at("n", ~.x / patients) %>%
  select(-patients) %>%
  pivot_wider(names_from = "type", values_from = "n") %>%
  filter(group != "(Missing)",
         pod_type != "Bed") %>%
  mutate_at("group",
            fct_recode,
            "OTI" = "Other Terminal Illness") %>%
  ggplot(aes(pod_type, stp)) +
  geom_col(aes(colour = group,
               fill = after_scale(alpha(colour, 0.4))),
           position = position_dodge(width = 0.8),
           width = 0.8) +
  geom_point(aes(y = region,
                 colour = group,
                 fill = after_scale(alpha(colour, 0.4))),
             show.legend = FALSE,
             shape = "circle filled",
             size = 2,
             stroke = 1.5,
             position = position_dodge(width = .8))+
  scale_y_continuous(labels = percent,
                     expand = expansion(c(0, 0.05))) +
  expand_limits(y = c(0, 1)) +
  labs(x = "", y = "", colour = "", fill = "") +
  theme(legend.position = "bottom",
        axis.line.x = element_blank(),
        axis.ticks.x = element_blank())
```

## Service use differs radically by cause of death

Behind headline levels of access examining patients’ interactions with services can reveal
significantly different patterns. These patterns can then suggest specific areas for improvement. To
illustrate this, we now investigate the service interactions of a small sample of fifteen random
decedents. We do so for each cause of death, focusing on their last year of life.

Figure 14 shows the format used for this analysis. It does so here using a ‘fictional’ example of a
decedent’s service use in the last year of life. The chart shows different service events, starting at
day 365 and ending at death. This format is applied below to look at samples of decedents for
different causes of death. 

#### Figure 14 : Fictional example explaining patterns of service use charts

*[insert figure 14]*

### Most of those dying from frailty access urgent care

Analysis of the sample of those dying from frailty (Figure 15) suggests that a large proportion of
people experience an emergency admission, plus an associated bed stay, at some point in their last
year. The likelihood of this increases the closer people are to death.

#### Figure 15 : Patterns of service use for people dying from frailty

*[insert figure 15]*

### Planned care features highly for those dying of cancer

Analysis of the sample of those dying from cancer (Figure 16) suggests frequent planned contacts
and planned admissions. This group is also more likely to have a planned stay in hospital and
experience more planned bed days than other cause of death groups. Urgent events and
associated urgent bed stays are more likely to occur in the last six months of life.

#### Figure 16 : Patterns of service use for people dying from cancer

*[insert figure 16]*

### People dying from organ failure experience multiple urgent stays

Figure 17 illustrates service use for people dying of organ failure. Here there are multiple urgent
events with an associated bed stay in the last year of life. The closer this population are to death,
the longer these stays tend to become.

#### Figure 17 : Patterns of service use for people dying from organ failure

*[insert figure 17]*

### People dying a sudden death tend to experience urgent care

Figure 18 shows the sample of people experiencing sudden death. Here there are planned contacts
throughout the last year of life. A large proportion experience an urgent event with associated
urgent bed stay in their final month - often then going on to die in hospital after a short period.

#### Figure 18 : Patterns of service use for people dying a sudden death

*[insert figure 18]*

### Those dying from other terminal illness experience long lengths of stay

Analysis for people dying of ‘other terminal illness’ (Figure 19) presents a mixed picture. A large
proportion experience an urgent service event with long bed stays. When charts show urgent bed
days intersected by another urgent event this represents patients being transferred between
different hospitals.

#### Figure 19 : Patterns of service use for people dying from other terminal illness

*[insert figure 19]*
