# How much is spent and what level of resource will be required in future?

In earlier section (section 6) we have examined use of services from the perspective of activity. Now
we consider use in terms of spend for `r stp_name` STP decedents in the
two years before they die. We then move on to consider the level of resources required by future
decedent populations from perspective of activity, spend and beds.

##  Urgent care accounts for two-thirds of expenditure

```{r 8.1 figures}
stp_cost <- activity_costs %>%
  filter(pod_type %in% c("Urgent Service Event",
                         "Planned Contact",
                         "Planned Admission",
                         "Critical Care")) %>%
  pull(cost) %>%
  sum()
avg_cost <- sum(activity_costs$cost) / nrow(mpi)
```

The calculated total hospital spend in the last two years of life in the `r stp_name`
is `r number(stp_cost, scale = 1e-6, accuracy = 1, prefix = "£")` million.
Figure 51 shows spend by activity type. Urgent services dominate  spend, consuming two-thirds of
end of life resource.

#### Figure 51 : Total spend by activity type in two years prior to death – `r stp_name` STP

```{r total spend by activity type (stp)}
activity_costs %>%
  filter(!pod_type %in% c("Bed", "Device", "Drug", "Other")) %>%
  group_by(pod_type) %>%
  summarise_at("cost", sum) %>%
  ggplot(aes(pod_type, cost)) +
  geom_col(aes(colour = pod_type,
               fill = after_scale(alpha(colour, 0.4))), show.legend = FALSE) +
  geom_text(aes(label = money_format(cost)),
                vjust = -0.35) +
  scale_y_continuous(labels = money_format,
                     expand = expansion(c(0, 0.05))) +
  labs(x = "", y = "")
```

## STP expenditure differs widely in the `r region_name`

The average spend in the last two years of life in the `r stp_name` is
`r comma(avg_cost, prefix = "£")` per decedent. To put this into context, government spend on
hospital care per year per head is £1,225.

Figure 52 shows for each STP average spend and confidence interval – the range in which we can
be reasonably confident that the true average lies. *[insert text]*.

#### Figure 52 : Average spend per decedent in two years prior to death – by STP with `r region_name` regional average and STP confidence intervals indicated by whiskers 

```{r average spend by stp}
mpi_region %>%
  left_join(activity_costs_region, by = "su_pat_id") %>%
  group_by(stp, su_pat_id) %>%
  summarise_at("cost", sum) %>%
  modify_at("cost", replace_na, 0) %>%
  summarise(ci = list(cost) %>% map(DescTools::MeanCI), .groups = "drop") %>%
  unnest_wider(ci) %>%
  inner_join(stps, by = c("stp" = "stp18cd")) %>%
  mutate_at("stp18nm", fct_reorder, quo(mean)) %>%
  mutate(colour = ifelse(stp == params$stp, "orange", "charcoal")) %>%
  mutate(region = sum(activity_costs_region$cost) / nrow(mpi_region)) %>%
  mutate(region_text = ifelse(mean == min(mean),
                              paste0(region_name, "\n",
                                     comma(region, prefix = "£", accuracy = 1)),
                              NA)) %>%
  ggplot(aes(mean, stp18nm)) +
  geom_col(aes(colour = colour, fill = after_scale(alpha(colour, 0.4))),
           show.legend = FALSE) +
  geom_errorbar(aes(xmin = lwr.ci, xmax = upr.ci), width = 0.5) +
  geom_text(aes(x = 0, label = comma(mean, prefix = "£", accuracy = 1)),
            hjust = -0.15) +
  geom_vline(aes(xintercept = region)) +
  geom_text(aes(x = region, label = region_text),
            na.rm = TRUE,
            size = 3,
            hjust = 0.6,
            vjust = -0.2,
            angle = 270) +
  scale_colour_manual(values = su_theme_cols()) +
  scale_x_continuous(labels = comma_format(prefix = "£", accuracy = 1)) +
  labs(x = "Average spend per Decedent", y = "") +
  theme(axis.line = element_blank(),
        axis.ticks.y = element_blank())
```

## Spend increases as death nears, but reduces in the final days

The spend curve by proximity to death shown in Figure 53 is similar to the utilisation curve of
urgent service events seen in sub-section 6.1.1. There is however one key difference noticeable
when we zoom in on the final month of life in Figure 54. This shows that for the `r region_name`, unlike
service use where activity peaks on day of death, spend peaks a few days before death. Activity still
takes place but the investigations, treatments and procedures which drive costs start to reduce.
Across the two years the rate and pattern of spend is generally the same as the `r region_name`. *[insert text]*.

#### Figure 53 : Average daily spend per decedent over two years - the `r stp_name` (yellow dots and line) and the `r region_name` region (grey line)

```{r average daily spend}
average_daily_spend_plots <- local({
  df <- list(
    stp = activity_costs,
    region = activity_costs_region
  ) %>%
    map(group_by, proximity_to_death_days) %>%
    map_dfr(summarise_at, "cost", sum, .id = "type") %>%
    pivot_wider(names_from = type, values_from = cost) %>%
    mutate_at("stp", `/`, nrow(mpi)) %>%
    mutate_at("region", `/`, nrow(mpi_region))
  
  c(Inf, 31) %>%
    map(function(.x) {
      p <- df %>%
        filter(proximity_to_death_days < .x) %>%
        ggplot(aes(proximity_to_death_days, stp)) +
        geom_point(colour = su_theme_cols("orange"),
                   size = 1,
                   fill = alpha(su_theme_cols("orange"), 0.4),
                   shape = "circle filled") +
        geom_smooth(method = "loess",
                    formula = y ~ x,
                    span = 0.2,
                    colour = su_theme_cols("orange"),
                    fill = NA) +
        geom_smooth(aes(y = region),
                    method = "loess",
                    formula = y ~ x,
                    span = 0.2,
                    colour = after_scale(alpha(su_theme_cols("charcoal"), 0.4)),
                    linetype = "solid",
                    fill = NA) +
        scale_y_continuous(labels = number_format(prefix = "£"),
                           position = "right") +
        theme(axis.title.y.right = element_text(margin = margin(l = 10))) +
        ylab("Average cost per decedent")
      
      if(is.finite(.x)) {
        p <- p +
          scale_x_continuous(trans = "reverse",
                             breaks = seq(0, 31, by = 7)) +
          coord_cartesian(xlim = c(.x, 0)) +
          xlab("Proximity to Death (Days)")
      } else {
        p <- p +
          scale_x_proximity_to_death_days() +
          xlab("Proximity to Death (Months)")
      }
      
      p
    })
})
average_daily_spend_plots[[1]]
```

#### Figure 54 : Average daily spend per decedent in final month - `r stp_name` STP (yellow dots and line) and the `r region_name` region (grey line)

```{r average daily spend (last month)}
average_daily_spend_plots[[2]]
```

## How will service use and expenditure evolve in future?

The earlier section (section 6) showed how much service use there was for those who died. Now we
consider how much will be used in future. We examine what levels of future use will be if current
patterns of service and resource use continue. We view these at five-year intervals and by age group
for activity, spend and beds.

Future levels of service use in `r stp_name` were predicted by applying
the current utilisation rates to future expected decedent populations. If utilisation rates were to
change then so too would the predicted levels. 

```{r forecast spend}
average_costs_per_person <- activity_costs %>%
  filter(pod_type != "Bed") %>%
  inner_join(mpi, by = "su_pat_id") %>%
  mutate_at("age", pmin, 90) %>%
  mutate(activity_year = as.numeric(proximity_to_death_days >= 365)) %>%
  group_by(activity_year, pod_type, age, sex) %>%
  summarise(cost = sum(cost), .groups = "drop") %>%
  complete(nesting(pod_type),
           activity_year = 0:1,
           age = seq(18, 90),
           sex,
           fill = list(cost = 0)) %>%
  inner_join(count(mpi, age, sex, name = "decedents"), by = c("age", "sex"))

average_costs_per_person_sum <- forecast_deaths %>%
  inner_join(average_costs_per_person, by = c("sex", "age_group" = "age")) %>%
  mutate_at("year", ~.x - activity_year) %>%
  filter(!year %% 5) %>%
  filter(age_group >= 18) %>%
  mutate_at("age_group",
            cut,
            c(0, 18, 65, 75, 85, Inf),
            labels = levels(mpi$age_band),
            right = FALSE) %>%
  group_by(year, age_group, pod_type) %>%
  summarise(across(c("est_deaths", "cost", "decedents"), sum), .groups = "drop_last")

forecast_spend_plots <- bind_rows(
  average_costs_per_person_sum,
  average_costs_per_person_sum %>%
    group_by(est_deaths, decedents, .add = TRUE) %>%
    summarise_at("cost", sum) %>%
    mutate(pod_type = "All Activity")
) %>%
  ungroup() %>%
  mutate_at("pod_type",
            fct_relevel,
            c("All Activity",
              levels(activity_costs$pod_type) %>% subset(., . != "Bed"))) %>%
  mutate(cost = ifelse(decedents, cost / decedents * est_deaths, 0)) %>%
  group_nest(pod_type) %>%
  mutate(plot = map(data, function(data) {
    lbls <- data %>%
      group_by(year) %>%
      summarise_at("cost", sum) %>%
      arrange(year) %>%
      mutate(act_sum_lbl = comma(cost, prefix = "£", accuracy = .1, scale = 1e-6, suffix = "m"),
             pcnt_change_lbl = percent(ifelse(year == 2020,
                                              NA,
                                              cost / first(cost) - 1),
                                       accuracy = .1))
  
    data %>%
      ggplot(aes(year, cost)) +
      geom_col(aes(fill = fct_rev(age_group)),
               alpha = 0.4,
               colour = "white",
               position = "stack") +
      geom_text(data = lbls, aes(label = act_sum_lbl), vjust = -.3, size = 3.5, na.rm = TRUE) +
      geom_text(data = lbls, aes(label = pcnt_change_lbl), vjust = 1.3, size = 3.5, na.rm = TRUE) +
      scale_fill_discrete(guide = guide_legend(reverse = TRUE),
                          reverse = TRUE) +
      scale_y_continuous(labels = money_format,
                         expand = expansion(c(0, 0.15))) +
      labs(fill = "", y = "Spend (£million)") +
      theme(legend.position = "bottom")
  }))

forecast_data <- forecast_activity %>%
  mutate_at("pod_type",
            ~ifelse(pod_summary_group == "Critical Care Bed Day",
                    "Critical Care Bed Day",
                    as.character(.x))) %>%
  mutate_at("pod_type",
            fct_relevel,
            c(levels(activity$pod_type), "Critical Care Bed Day")) %>%
  filter(year %% 5 == 0) %>%
  group_by(age_band, year, pod_type) %>%
  summarise_at(vars(est_deaths, activity), sum) %>%
  ungroup()

forecast_activity_plots <- forecast_data %>%
  group_nest(pod_type) %>%
  mutate(plot = map(data, function(data) {
    lbls <- data %>%
      group_by(year) %>%
      summarise_at("activity", sum) %>%
      arrange(year) %>%
      mutate(act_sum_lbl = comma(activity, accuracy = 1),
             pcnt_change_lbl = percent(ifelse(year == 2020,
                                              NA,
                                              activity / first(activity) - 1),
                                       accuracy = .1))
    
    data %>%
      ggplot(aes(year, activity)) +
      geom_col(aes(fill = fct_rev(age_band)),
               alpha = 0.4,
               colour = "white",
               position = "stack") +
      geom_text(data = lbls, aes(label = act_sum_lbl), vjust = -.3, na.rm = TRUE) +
      geom_text(data = lbls, aes(label = pcnt_change_lbl), vjust = 1.3, size = 3.5, na.rm = TRUE) +
      scale_y_continuous(labels = scales::comma_format(scale = .001),
                         expand = expansion(c(0, 0.15))) +
      scale_fill_discrete(guide = guide_legend(reverse = TRUE),
                          reverse = TRUE) +
      labs(fill = "",
           y = "Activity (1,000's)") +
      theme(legend.position = "bottom")
  }))
```

#### Urgent services activity and expenditure is set to increase

By 2030 urgent service events will have increased by *[insert %]* for activity (Figure 55) and *[insert %]* for spend
(Figure 56). Older decedents (those aged 85+) are the largest share of urgent services and growth
is driven by increasing number of decedents in this age group.

#### Figure 55 : Urgent service events future use at five-year intervals by decedent age group (percentages relative to 2020)

```{r forecast activity - urgent service event}
filter(forecast_activity_plots, pod_type == "Urgent Service Event")$plot[[1]]
```

#### Figure 56 : Urgent service events future spend at five-year intervals by decedent age group (percentages relative to 2020)

```{r forecast spend - urgent service event}
filter(forecast_spend_plots, pod_type == "Urgent Service Event")$plot[[1]]
```

### Future use and spend grow steadily for planned contacts

Jointly the two oldest age groups (those 75-84 and those aged 85+) are responsible for over half of
planned contacts. For both activity (Figure 57) and spend (Figure 58). Growth is driven by
increasing number of decedents in the oldest age group.

#### Figure 57 : Planned contact future use at five-year intervals by decedent age group (percentages relative to 2020)

```{r forecast activity - planned contacts}
filter(forecast_activity_plots, pod_type == "Planned Contact")$plot[[1]]
```

#### Figure 58 : Planned contact future spend at five-year intervals by decedent age group (percentages relative to 2020)

```{r forecast spend - planned contacts}
filter(forecast_spend_plots, pod_type == "Planned Contact")$plot[[1]]
```

### Future use and spend remain steady for planned admissions

Future levels of planned admissions remain steady for activity (Figure 59) and grow slowly for
spend (Figure 60). The increase in spend is proportionally more than the increase in activity.
Although younger age groups feature more heavily in planned admissions the growth in spend is
driven by decedents in the oldest age group.

#### Figure 59 : Planned admission future use at five-year intervals by decedent age group (percentages relative to 2020)

```{r forecast activity - planned admission}
filter(forecast_activity_plots, pod_type == "Planned Admission")$plot[[1]]
```

#### Figure 60 : Planned admission future spend at five-year intervals by decedent age group (percentages relative to 2020)

```{r forecast spend - planned admission}
filter(forecast_spend_plots, pod_type == "Planned Admission")$plot[[1]]
```

### Increased demand for bed days will put pressure on capacity

Future bed days have similar growth to urgent service events. By 2030 they will have increased by
*[insert %]* (Figure 61). Older decedents (those aged 85+) are responsible for the largest share of bed days
and growth is also driven by the increasing number of decedents in this age group. By 2030 to
meet growth in bed days `r stp_name` will require an additional *[insert #]* beds (Table 3).

Costs for future bed days are not calculated separately. They are included in cost of admission.

#### Figure 61 : Bed day future level of use at five-year intervals by decedent age group (percentages relative to 2020)

```{r forecast activity - bed}
filter(forecast_activity_plots, pod_type == "Bed")$plot[[1]]
```

#### Table 3 : Additional beds required to meet future growth in bed days

```{r forecast bed table}
forecast_data %>%
  filter(pod_type == "Bed") %>%
  group_by(year) %>%
  summarise(across(activity, sum), .groups = "drop") %>%
  mutate(across(activity, ~round((.x - first(.x)) / (365*.9)))) %>%
  filter(year > 2020) %>%
  pivot_wider(names_from = year, values_from = activity) %>%
  mutate(year = "Extra Beds") %>%
  relocate(year, .before = everything())
```

### Future use and spend reduces for critical care

Future level of critical care bed days reduce for both days (Figure 62) and spend (Figure 63). The
youngest group in the decedent population (aged 18-64) have the largest share of critical care
days. The reduction is driven by the decreasing number of decedents in this age group.

#### Figure 62 : Critical care bed day future use at five-year intervals by decedent age group (percentages relative to 2020)

```{r forecast activity - critical care bed day}
filter(forecast_activity_plots, pod_type == "Critical Care Bed Day")$plot[[1]]
```

#### Critical care bed day future spend at five-year intervals by decedent age group (percentages relative to 2020)

```{r forecast spend - critical care bed day}
filter(forecast_spend_plots, pod_type == "Critical Care")$plot[[1]]
```
