# Is there evidence of non-beneficial treatment in the final few weeks of life?

For those who are dying time is an increasingly precious and scarce resource. What happens in the
final few weeks of life is of paramount importance. In this section we consider non-beneficial
treatments in the final few weeks of life. A systematic review of non-beneficial treatments,
describes these as “a treatment that was administered with little or no hope of it having an effect,
largely because of the underlying state of the patient’s health and the known or expected poor
prognosis regardless of treatment”. In this section we consider non-beneficial treatments in the
following two areas:

- chemotherapy in the period four weeks prior to death
- use of critical care by the decedent population

## Is there evidence of non-beneficial use of chemotherapy?

Chemotherapy overuse close to the time of death has been suggested as a potential indicator of
poor quality of care. Additionally, a recent study found that palliative chemotherapy can in some
cases both shorten and reduce quality of life.

In the following sub-sections we consider chemotherapy decedents (those decedents who received
chemotherapy treatment in the two years prior to death). We consider how many received
chemotherapy close to death, where ‘close to death’ is defined as the period four weeks prior to
death. Moving on to consider patterns of chemotherapy service use and start date of treatment
before investigating differences by demographic and clinical subgroups.

### One in six of those receiving chemotherapy did so in the last month of life

```{r 7.1.1 figures}
chemo_last_month <- mpi %>%
  inner_join(filter(activity, chemotherapy_indicator == 1), by = "su_pat_id") %>%
  group_by(su_pat_id) %>%
  summarise(last_month = any(proximity_to_death_days <= 28), .groups = "drop") %>%
  summarise(n = n(), across(last_month, sum), p = last_month / n)
```

Of `r comma(chemo_last_month$n)` chemotherapy decedents in the `r stp_name` `r comma(chemo_last_month$last_month)`
(`r percent(chemo_last_month$p, accuracy = 1)`) received chemotherapy in the last four weeks of life.
The remaining `r comma(chemo_last_month$n - chemo_last_month$last_month)`
(`r percent(1 - chemo_last_month$p, accuracy = 1)`) did not receive chemotherapy
their last four weeks. Characteristics of these two groups are compared in the following subsections.

### Pattern of service use differs for those who have chemotherapy close to death

Patterns differ significantly between the two groups. Figure 38 shows that for much of the two
years those receiving chemotherapy in the last four weeks of life use comparatively less
chemotherapy than those who do not. Their use of chemotherapy starts to increase more rapidly
around the same time that use by the other group declines. It then peaks close to death. This is the
same pattern as seen in the `r region_name` (Figure 39).

#### Figure 38 : Chemotherapy use per 1,000 chemotherapy decedents by proximity to death in days - `r stp_name` STP

```{r chemo activity split last month plot (stp)}
chemo_activity_split_plot <- function(activity) {
  activity %>%
    filter(chemotherapy_indicator == 1) %>%
    group_by(su_pat_id) %>%
    mutate(last_month = any(proximity_to_death_days <= 28)) %>%
    group_by(last_month) %>%
    mutate(nd = n_distinct(su_pat_id)) %>%
    count(proximity_to_death_days, nd) %>%
    ungroup() %>%
    mutate_at("nd", ~n / nd) %>%
    complete(last_month, proximity_to_death_days, fill = list(nd = 0, n = 0)) %>%
    filter(last_month | proximity_to_death_days > 28)  %>%
    mutate_at("last_month",
              ifelse,
              "later - received in last 4 weeks",
              "earlier - did not receive in last 4 weeks") %>%
    ggplot(aes(proximity_to_death_days, nd, colour = last_month)) +
    geom_smooth(fill = NA,
                method = "loess",
                formula = y ~ x) +
    geom_vline(xintercept = 28, linetype = "dashed") +
    scale_x_proximity_to_death_days() +
    scale_y_continuous(labels = comma_format(scale = 1000),
                       position = "right") +
    theme(legend.position = "bottom",
          axis.title.y.right = element_text(margin = margin(0, 0, 0, 10)),
          panel.grid.major = element_line(linetype = "dotted",
                                          colour = "lightgrey")) +
    labs(x = "Proximity to death (Months)",
         y = "Daily chemo activity\nper 1,000 decedents who had chemo",
         colour = "") +
    annotate("text",
             x = 28,
             y = 0,
             label = "4 weeks prior to death",
             hjust = 1.05,
             vjust = -0.5,
             size = 3)
}

chemo_activity_split_plot(activity)
```

#### Figure 39 : Chemotherapy use per 1,000 chemotherapy decedents by proximity to death in days - `r region_name` region

```{r chemo activity split last month plot (region)}
chemo_activity_split_plot(activity_region)
```

### People having chemotherapy close to death start treatment later

```{r chemo when do people have first chemo split last month}
first_chemo_split_plot <- function(activity) {
  activity %>%
    filter(chemotherapy_indicator == 1) %>%
    group_by(su_pat_id) %>%
    summarise(last_month = any(proximity_to_death_days <= 28),
              proximity_to_death_days = max(proximity_to_death_days),
              .groups = "drop_last") %>%
    count(proximity_to_death_days, last_month) %>%
    complete(proximity_to_death_days, last_month, fill = list(n = 0)) %>%
    filter(last_month | proximity_to_death_days > 28) %>%
    mutate_at("last_month",
              ifelse,
              "later - received in last 4 weeks",
              "earlier - did not receive in last 4 weeks") %>%
    group_by(last_month) %>%
    arrange(desc(proximity_to_death_days)) %>%
    mutate_at("n", cumsum) %>%
    mutate(p = n / max(n)) %>%
    # there is a slight dip in the first couple of days of data - this "cleans"
    # the chart up a little
    filter(proximity_to_death_days < 365*1.95) %>%
    ggplot(aes(proximity_to_death_days, p, colour = last_month)) +
    geom_step(lwd = 1) +
    geom_vline(xintercept = 28, linetype = "dashed") +
    scale_y_continuous(labels = percent,
                       position = "right") +
    scale_x_proximity_to_death_days() +
    labs(x = "Proximity to death (Months)",
         y = "Cumulative % of when people had\ntheir first chemo",
         colour = "") +
    theme(legend.position = "bottom",
          panel.grid.major = element_line(linetype = "dashed",
                                          colour = "lightgrey"),
          axis.title.y.right = element_text(margin = margin(l = 10))) +
    annotate("text",
             x = 28,
             y = 0,
             label = "4 weeks prior to death",
             hjust = 1.05,
             vjust = -0.5,
             size = 3)
}
```

Those receiving chemotherapy in the last four weeks of life start chemotherapy much later. Figure
40 shows the cumulative proportion of chemotherapy decedents by start of chemotherapy date. *[insert text]*. 

#### Figure 40 : Cumulative proportion of chemotherapy decedents by start date - `r stp_name` STP

```{r chemo when do people have first chemo split last month (stp)}
first_chemo_split_plot(activity)
```

#### Figure 41 : Cumulative proportion of chemotherapy decedents by start date - `r region_name` region

```{r chemo when do people have first chemo split last month (region)}
first_chemo_split_plot(activity_region)
```

### People having late chemotherapy tend to be younger

Figure 42 shows composition by age group of those in `r stp_name`
receiving chemotherapy in the last four weeks of life. *[insert text]*.

```{r chemo in last month of life binomial ci plot by age band}
chemo_last_month_binomial_ci_plot_age <- function(activity) {
  activity %>%
    filter(chemotherapy_indicator == 1) %>%
    inner_join(select(mpi_region, su_pat_id, age_band), by = "su_pat_id") %>%
    mutate(last_month = ifelse(proximity_to_death_days <= 28, su_pat_id, NA)) %>%
    group_by(age_band) %>%
    summarise(d = n_distinct(su_pat_id),
              n = n_distinct(last_month, na.rm = TRUE),
              .groups = "drop_last") %>%
    mutate(ci = map2(n, d, compose(as_tibble, BinomCI))) %>%
    unnest(ci) %>%
    mutate(mean_est = sum(n) / sum(d),
           colour = case_when(
             upr.ci < mean_est ~ "significantly lower",
             lwr.ci > mean_est ~ "significantly higher",
             TRUE ~ "no significant difference",
           ) %>%
             factor(levels = c("significantly lower",
                               "no significant difference",
                               "significantly higher"))) %>%
    ggplot(aes(fct_rev(age_band), est, colour = colour)) +
    geom_errorbar(aes(ymin = lwr.ci, ymax = upr.ci)) +
    geom_point(aes(fill = after_scale(alpha(colour, 0.4)))) +
    geom_hline(aes(yintercept = mean_est),
               linetype = "dashed") +
    coord_flip() +
    scale_colour_discrete(drop = FALSE,
                          guide = guide_legend(reverse = TRUE)) +
    scale_y_continuous(labels = percent_format(accuracy = 1)) +
    labs(x = "",
         y = paste("% of patients who have chemo in last month before death",
                   "Dashed line = average %",
                   sep = "\n")) +
    theme(axis.line.y = element_blank(),
          axis.ticks.y = element_blank(),
          panel.grid.major.x = element_line(colour = "lightgrey",
                                            linetype = "dotted"),
          legend.position = c(0, 1),
          legend.justification = c(0, 1),
          legend.background = element_blank(),
          legend.title = element_blank())
}
```

#### Figure 42 : Proportion by age group receiving chemotherapy in the last four weeks of life - `r stp_name` STP

```{r chemo in last month of life binomial ci plot by age band (stp)}
chemo_last_month_binomial_ci_plot_age(activity)
```

#### Figure 43 : Proportion by age group receiving chemotherapy in the last four weeks of life - `r region_name` region

```{r chemo in last month of life binomial ci plot by age band (region)}
chemo_last_month_binomial_ci_plot_age(activity_region)
```

### People having late chemotherapy have certain types of cancer

Figure 44 shows composition by cancer type of those in the `r stp_name`
receiving chemotherapy in the last four weeks of life. *[insert text]*

```{r chemo in last month of life binomial ci plot by cwt site}
cwt_groups <- file.path("data","reference","cwt_groups.csv") %>%
  read_csv(col_types = "cc")

chemo_last_month_binomial_ci_plot_cwt <- function(activity) {
  activity %>%
    filter(chemotherapy_indicator == 1) %>%
    inner_join(select(mpi_region, su_pat_id, primary_cod), by = "su_pat_id") %>%
    mutate_at("primary_cod", ~ifelse(.x == "C770", .x, str_sub(.x, 1, 3))) %>%
    inner_join(cwt_groups, by = c("primary_cod" = "icd10")) %>%
    mutate(last_month = ifelse(proximity_to_death_days <= 28, su_pat_id, NA)) %>%
    group_by(cancer_site) %>%
    summarise(d = n_distinct(su_pat_id),
              n = n_distinct(last_month, na.rm = TRUE),
              .groups = "drop_last") %>%
    mutate(ci = map2(n, d, compose(as_tibble, BinomCI))) %>%
    unnest(ci) %>%
    mutate(mean_est = sum(n) / sum(d),
           colour = case_when(
             upr.ci < mean_est ~ "significantly lower",
             lwr.ci > mean_est ~ "significantly higher",
             TRUE ~ "no significant difference",
           ) %>%
             factor(levels = c("significantly lower",
                               "no significant difference",
                               "significantly higher"))) %>%
    mutate_at("cancer_site", fct_reorder, quo(-est)) %>%
    ggplot(aes(fct_rev(cancer_site), est, colour = colour)) +
    geom_errorbar(aes(ymin = lwr.ci, ymax = upr.ci)) +
    geom_point(aes(fill = after_scale(alpha(colour, 0.4)))) +
    geom_hline(aes(yintercept = mean_est),
               linetype = "dashed") +
    coord_flip() +
    scale_colour_discrete(drop = FALSE,
                          guide = guide_legend(reverse = TRUE)) +
    scale_y_continuous(labels = percent_format(accuracy = 1)) +
    labs(x = "",
         y = paste("% of patients who have chemo in last month before death",
                   "Dashed line = average %",
                   sep = "\n")) +
    theme(axis.line.y = element_blank(),
          axis.ticks.y = element_blank(),
          panel.grid.major.x = element_line(colour = "lightgrey",
                                            linetype = "dotted"),
          legend.position = c(1, 0),
          legend.justification = c(1, 0),
          legend.background = element_blank(),
          legend.title = element_blank())
}
```

#### Figure 44 : Proportion by cancer type receiving chemotherapy in the last four weeks of life - `r stp_name` STP

```{r chemo in last month of life binomial ci plot by cwt site (stp)}
chemo_last_month_binomial_ci_plot_cwt(activity)
```

#### Figure 45 : Proportion by cancer type receiving chemotherapy in the last four weeks of life - `r region_name` region

```{r chemo in last month of life binomial ci plot by cwt site (region)}
chemo_last_month_binomial_ci_plot_cwt(activity_region)
```

## How is critical care used at end of life?

Critical care units are crucially important, they are highly specialist wards that provide treatment
and monitoring for people who become very unwell or are recovering from major surgery. 1 in
every 8 decedents (sub-section 5.1) spend some time in critical care in the two years before they
die. For some decedents, particularly when close to death, critical care may be non-beneficial. It is
also a costly resource. This is important because in addition to a lack of benefit to patients use of
critical care also has substantial resource implications for services.

We now consider critical care in detail. We investigate how days in critical care are used by
decedents and for how many decedents it is their place of death.

### Critical care bed days are usually part of emergency care

Critical care is usually a component of a longer hospital stay. Some decedents who spend time in
this setting do so after elective surgery whilst others do so following emergency admission. For
those decedents using critical care Figure 46 shows the amount of critical care days by source. It
shows that:

- More days are used by decedents with organ failure (1 in 3 of all critical care bed days) than
other causes; and,
- Most critical care bed days used by the decedent population are part of an emergency
admission. Critical care days after elective admission are few but cancer is an exception, close
to a third of critical care days for cancer decedents are part of an elective admission.
- When compared to the `r region_name` (Figure 47) there are more critical care days used by organ
failure decedents. Profiles by type of admission are similar across causes of death.

#### Figure 46 : Proportion of critical care bed days by cause and source of admission - `r stp_name` stp

```{r critical care bed days - admission source (stp)}
activity %>%
  inner_join(select(mpi, su_pat_id, group),
             by = "su_pat_id") %>%
  filter(group != "(Missing)",
         pod_summary_group == "Critical Care Bed Day") %>%
  distinct(bb5008_pseudo_apcs_ident_pseudo, group, cc_pod_summary_group) %>%
  mutate_at(vars(group),
            compose(fct_rev, fct_recode),
            "Sudden\nDeath" = "Sudden Death",
            "OTI" = "Other Terminal Illness") %>%
  mutate_at("cc_pod_summary_group", str_replace, " ", "\n") %>%
    mekko_chart(group, cc_pod_summary_group)
```

#### Figure 47 : Propotion of critical care bed days by cause and source of admission - `r region_name` region

```{r critical care bed days - admission source (region)}
activity_region %>%
  inner_join(select(mpi_region, su_pat_id, group),
             by = "su_pat_id") %>%
  filter(group != "(Missing)",
         pod_summary_group == "Critical Care Bed Day") %>%
  distinct(bb5008_pseudo_apcs_ident_pseudo, group, cc_pod_summary_group) %>%
  mutate_at(vars(group),
            compose(fct_rev, fct_recode),
            "Sudden\nDeath" = "Sudden Death",
            "OTI" = "Other Terminal Illness") %>%
  mutate_at("cc_pod_summary_group", str_replace, " ", "\n") %>%
    mekko_chart(group, cc_pod_summary_group)
```

### Death in critical care varies greatly by cause of death

Critical care stays end when a patient leaves critical care, either because they have recovered
enough to be moved to another ward or because they have died in critical care. Figure 48 shows
the amount of days spent in critical care by the eventual outcomes of:

- Did not die in critical care – moved to non-critical care ward; or,
- Died in critical care.

It shows that for sudden death more than half of critical care days are for a stay which ends with
death in critical care. The proportions are similarly high for other terminal illness and organ failure.
Cancer has the smallest proportion of days in critical care ending in death. When compared to the
`r region_name` (Figure 49) profiles of critical care days by cause are similar.

#### Figure 48 : Proportion of critical care bed days by cause and outcome - `r stp_name` STP

```{r critical care bed days - outcome (stp)}
mpi %>%
  filter(group != "(Missing)") %>%
  mutate_at(vars(group),
            compose(fct_rev, fct_recode),
            "Sudden\nDeath" = "Sudden Death",
            "OTI" = "Other Terminal Illness") %>%
  drop_na(died_in_critical_care) %>%
  mutate_at("died_in_critical_care", ~ifelse(.x == 1,
                                             "Died in\ncritical care",
                                             "Did not die in\ncritical care")) %>%
  mekko_chart(group, died_in_critical_care)
```

#### Figure 49 : Proportion of critical care bed days by cause and outcome - `r region_name` region

```{r critical care bed days - outcome (region)}
mpi_region %>%
  filter(group != "(Missing)") %>%
  mutate_at(vars(group),
            compose(fct_rev, fct_recode),
            "Sudden\nDeath" = "Sudden Death",
            "OTI" = "Other Terminal Illness") %>%
  drop_na(died_in_critical_care) %>%
  mutate_at("died_in_critical_care", ~ifelse(.x == 1,
                                             "Died in\ncritical care",
                                             "Did not die in\ncritical care")) %>%
  mekko_chart(group, died_in_critical_care)
```

### Fewer frailty and cancer deaths happen in critical care

We now consider stays, rather than days, in critical care. We investigate the proportion of critical
care stays ending in death as a proportion of all decedent critical care stays. Figure 50 shows the
proportions by cause. `r stp_name` are shown as larger yellow dots
(with confidence interval – the range in which we can be reasonably confident that the true
proportion lies); the `r region_name` is shown as smaller grey dots. It can be seen that:

- Proportionally fewer critical care stays end in death for cancer and frailty patients. This reflects
the nature of cancer and frailty patients who, with advanced incurable disease or pre-existing
limitations of treatment, are less likely to benefit from critical care;
- Proportionally more critical care stays end in death for the causes of other terminal illness,
sudden death and organ failure; and,
- The `r region_name` have fewer critical care stays ending in death for other terminal illness, cancer
and frailty. However, these are within confidence intervals and are therefore not considered
significantly different.

#### Figure 50 : Proportion of critical care spells ending in death, confidence intervals indicated by whiskers – `r stp_name` (yellow dots), `r region_name` (grey dots)

```{r critical care spells ending in death}
pcnt_dot_comparison_plot <- function(mpi, mpi_region, category, value, reorder = TRUE) {
  data <- list(
    stp = mpi,
    region = mpi_region
  ) %>%
    map(filter, {{category}} != "(Missing)", !is.na({{category}})) %>%
    map(group_by, {{category}}) %>%
    map_dfr(summarise,
            numerator = sum({{value}}, na.rm = TRUE),
            denominator = n(),
            .groups = "drop",
            .id = "type") %>%
    mutate(across(type, fct_relevel, "stp", "region"),
           x = map2(numerator, denominator, compose(as_tibble, BinomCI))) %>%
    unnest_wider(x) %>%
    rename({{value}} := est) %>%
    mutate_at(vars(lwr.ci, upr.ci), ~ifelse(type == "region", NA, .x))

  
  if (reorder) {
    data <- mutate(data, across({{category}}, fct_reorder, {{value}}))
  }
  
  ggplot(data, aes({{category}}, {{value}}, fill = type)) +
    geom_errorbar(aes(ymin = lwr.ci, ymax = upr.ci),
                  na.rm = TRUE,
                  width = 0.2) +
    geom_point(aes(size = type), shape = "circle filled") +
    coord_flip() +
    scale_size_manual(values = c(5, 2)) +
    scale_fill_manual(values = su_theme_cols()[c("orange", "slate")] %>%
                        unname()) +
    scale_y_continuous(labels = percent_format(accuracy = 1),
                       expand = expansion(mult = c(0, 0.05))) +
    expand_limits(y = 0) +
    theme(axis.line = element_blank(),
          axis.ticks.y = element_blank(),
          panel.grid.major.x = element_line(colour = "grey",
                                            linetype = "dotted"),
          legend.position = c(1, 0),
          legend.justification = c(1, 0),
          legend.direction = "vertical") +
    labs(x = "", y = "", fill = "", size = "")
}

pcnt_dot_comparison_plot(mpi, mpi_region, group, died_in_critical_care) +
  ylab("% who died in critical care")
```
